Grocery Pricing Strategies: An Analysis of Vendor Positioning and Market Dynamics

### 2024-11-14 paper/paper.qmd
polish paper

---
title: "Grocery Pricing Strategies: An Analysis of Vendor Positioning and Market Dynamics"
subtitle: "Examining Sales Frequency, Price Distribution, and Average Price Levels Across Major Grocery Chains"
author: 
  - Andy Jiang
thanks: "Code and data are available at: [https://github.com/AndyYanxunJiang/grocery-price-analysis.git)."
date: today
date-format: long
abstract: "This paper examines grocery pricing strategies across major vendors to understand how they position themselves in a competitive market. By analyzing sales frequency, price distribution, and average price levels, we identify distinct approaches used by each vendor to attract customers. Our findings reveal that some vendors, like Metro, rely on frequent promotions to appeal to a broad customer base, while others, such as No Frills and Walmart, prioritize everyday low pricing to cater to budget-conscious shoppers. This analysis highlights the varied tactics grocery vendors use to target different consumer segments, providing insights into the competitive dynamics that shape grocery pricing and customer choices."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

# Data {#sec-data}
We use the statistical programming language R [@citeR] and several libraries, including tidyverse [@tidyverse], DBI [@citeDBI], RSQLite [@citeRSQLite], and ggplot2 [@citeGG]. The data was sourced from Project Hammer [@Filipp_2024], a project database created by Jacob Filipp that compiles grocery price data from various vendors.

## Measurement
The dataset used in this analysis captures real-world grocery pricing and promotional strategies from multiple vendors. Each entry, sourced from Project Hammer [@Filipp_2024], represents a product listing, with variables such as current_price, old_price, vendor, and nowtime. These fields translate vendor pricing strategies into data points: current_price reflects the active selling price, while old_price denotes any discount or promotional pricing. The vendor field identifies the grocery chain, and nowtime timestamps each entry, allowing us to analyze trends over time. This structure enables an organized examination of pricing tactics across vendors by converting complex market behaviors into measurable data.

@@

# Discussion {#sec-discussion}
The results of this analysis offer valuable insights into the pricing and promotional strategies of various grocery vendors. However, several methodological considerations need to be addressed to ensure an accurate interpretation of the data. Understanding the distinction between correlation and causation, handling missing data, and accounting for potential sources of bias are essential to refining the conclusions drawn from this study. Each of these factors is discussed in the following subsections.

## Correlation vs. Causation
In analyzing grocery pricing and promotional strategies, it’s important to distinguish between correlation and causation. While our data reveals correlations—such as a relationship between high-frequency sales and lower average prices in certain vendors—these associations do not imply causation. For example, Metro’s high sales frequency correlates with its broader price range, but this does not necessarily mean that frequent sales cause price diversity, or vice versa. Factors like target market segmentation, inventory management, and strategic pricing decisions can all influence these metrics independently. Therefore, while the analysis highlights patterns in how different vendors structure pricing and promotions, it does not establish causative relationships between these variables.

## Missing Data
The dataset may contain missing values, particularly in fields such as current and old prices. These missing values can affect the accuracy of the analysis, especially in sections where price distribution and averages are calculated. For example, if a vendor has missing values during high promotional periods, their frequency of sales may be underestimated. Similarly, missing prices for premium products could bias the average price calculation downward. To mitigate this, we addressed missing data by filtering out incomplete rows for critical analyses; however, the absence of data for certain time periods or products still limits the representativeness of our findings.

## Sources of Bias
Several potential sources of bias could affect the analysis and interpretation of grocery pricing data. First, the data is derived from a limited selection of vendors, which may not represent all grocery stores in the region, particularly smaller, independent stores. Second, the data is based on specific geographic regions and may not account for regional pricing variations. Finally, since the data was collected from website listings, it may not reflect in-store promotions or exclusive discounts for loyalty program members, thus potentially underestimating actual promotional frequencies. These sources of bias should be considered when interpreting the results, as they may influence conclusions about vendor strategies and market positioning.

>>>>

---
title: "Grocery Pricing Strategies: An Analysis of Vendor Positioning and Market Dynamics"
subtitle: "Examining Sales Frequency, Price Distribution, and Average Price Levels Across Major Grocery Chains"
author: 
  - Andy Jiang
thanks: "Code and data are available at: [https://github.com/AndyYanxunJiang/grocery-price-analysis.git)."
date: today
date-format: long
abstract: "This paper examines grocery pricing strategies across major vendors to understand how they position themselves in a competitive market. By analyzing sales frequency, price distribution, and average price levels, we identify distinct approaches used by each vendor to attract customers. Our findings reveal that some vendors, like Metro, rely on frequent promotions to appeal to a broad customer base, while others, such as No Frills and Walmart, prioritize everyday low pricing to cater to budget-conscious shoppers. This analysis highlights the varied tactics grocery vendors use to target different consumer segments, providing insights into the competitive dynamics that shape grocery pricing and customer choices."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
---

@@

# Data {#sec-data}
We use the statistical programming language R [@citeR] and several libraries, including `tidyverse` [@tidyverse], `DBI` [@citeDBI], `RSQLite` [@citeRSQLite], and `ggplot2` [@citeGG]. The data was sourced from Project Hammer [@Filipp_2024], a project database created by Jacob Filipp that compiles grocery price data from various vendors.

## Measurement
The dataset used in this analysis captures real-world grocery pricing and promotional strategies from multiple vendors. Each entry, sourced from Project Hammer [@Filipp_2024], represents a product listing, with variables such as `current_price`, `old_price`, `vendor`, and `nowtime`. These fields translate vendor pricing strategies into data points: `current_price` reflects the active selling price, while `old_price` denotes any discount or promotional pricing. The vendor field identifies the grocery chain, and `nowtime` timestamps each entry, allowing us to analyze trends over time. This structure enables an organized examination of pricing tactics across vendors by converting complex market behaviors into measurable data.

@@

# Discussion {#sec-discussion}
The results of this analysis provide valuable insights into the pricing and promotional strategies of various grocery vendors, revealing distinct approaches across vendors like Metro, No Frills, Walmart, and others. However, to interpret these findings accurately, it is essential to address certain methodological considerations. These include distinguishing between correlation and causation, handling missing data, and recognizing potential sources of bias. These factors are critical to refining the conclusions drawn from the study and are discussed in detail below.

## Correlation vs. Causation
Our analysis revealed interesting correlations, such as Metro’s high frequency of sales being associated with a broad price range, suggesting a dual approach that includes both budget and premium offerings. Additionally, No Frills and Walmart showed consistently low prices and lower sales frequency, which aligns with a strategy that focuses on everyday low pricing. However, while these correlations are clear, they do not imply causation. For instance, Metro’s high sales frequency and diverse price range do not necessarily mean that frequent sales lead to a broader price distribution or that a wider range causes more frequent promotions. Both metrics could be influenced by other factors, such as Metro’s target market, inventory diversity, or management’s pricing strategy, which may independently affect both the frequency of sales and price diversity.

Similarly, vendors like No Frills and Walmart, with lower prices and fewer promotional events, might reflect a commitment to stable, low-cost offerings rather than suggesting that low prices cause a lower sales frequency. These observed patterns highlight potential correlations in vendor strategies but do not establish a causal link. Further research with controlled variables would be required to determine whether these strategies actively drive changes in other metrics or if they are simply co-occurring features of each vendor’s broader approach.

## Missing Data
The dataset contained missing values in key fields, such as `current_price` and `old_price`, which affected parts of the analysis, particularly in calculating accurate price averages and sales frequency. For instance, missing price data during high promotional periods for certain vendors could lead to an underestimation of their true sales frequency. This is particularly relevant for vendors like Metro, where sales frequency was a significant observation. If promotional pricing data was incomplete, Metro’s high frequency of sales might appear inflated or underestimated, depending on when the data was missing. Similarly, missing data in the `current_price` field for premium items could cause an underrepresentation of higher prices in the dataset, potentially leading to a lower calculated average price than is accurate. This would especially impact vendors with premium offerings, like Galleria, where higher average prices were noted.

To mitigate the effects of missing data, we filtered out incomplete rows in critical analyses, ensuring that results were based only on fully recorded entries. However, this approach cannot account for the representativeness of missing data over time and across vendors, and therefore, the conclusions drawn may be limited by these gaps. Addressing missing data remains essential for improving the accuracy of this type of analysis in future research.

## Sources of Bias
The dataset is subject to several potential biases that may affect the analysis and interpretation of grocery pricing data. First, the data includes only a select number of vendors, possibly excluding smaller, independent stores that may have different pricing and promotional strategies. This limitation could skew our understanding of average pricing, as larger vendors with more data points (e.g., Metro, Walmart) might dominate the analysis, potentially overshadowing strategies that smaller vendors might use. Additionally, the data is region-specific and does not account for regional variations in grocery pricing, meaning that the findings may not fully represent pricing trends across different areas.

Another important source of bias stems from the method of data collection. The data was gathered from online listings, which may differ from in-store pricing and promotions. For instance, in-store promotions or discounts exclusive to loyalty program members would not be captured, potentially underestimating the frequency and diversity of promotions offered by each vendor. This could particularly affect vendors like No Frills and Walmart, which tend to have stable, low prices and may offer in-store promotions not reflected online. Consequently, this method of data collection might result in an incomplete picture of vendor strategies, especially for promotional events and discounts.

These biases should be carefully considered when interpreting the results, as they may influence conclusions about vendor positioning and market strategy. While this analysis highlights general trends across major vendors, understanding the full scope of grocery pricing dynamics would require addressing these biases in future research, potentially by expanding the dataset to include more vendors, regions, and in-store data.
